### 2025-06-22 lecturenotes_aima104/ch3_aima104.tex
innerprodut

\subsection{Multiplicities of an eigenvalue}
\begin{itemize}
\item The algebraic multiplicity of $\lambda$ is the number of times $(\lambda - t)$ occurs as a factor of in its characteristic polynomial.
\item The geometric multiplicity of an eigenvalue $\lambda$ is the dimension of the corresponding eigenspace $E_{\lambda}$. i.e $dim\{E_{\lambda}\}$.
\end{itemize}
\subsection{Exercise}
Diagonalizable the following matrix, if possible.
$$A=
\begin{bmatrix}
  4 & 0 & 0  \\
  1 & 4 & 0  \\
   0& 0 & 5  \\
\end{bmatrix}
$$

@@

\section{Linear Transformation and Diagonalization}
We will see that the transformation $x \mapsto As$ is essentially same as the very simple mapping $u \maps to Du$ when viewed from the proper perspective.
\begin{theorem}
  Suppose $A = PDP^{-1}$, where $D$ is a diagonal $n \times n$. If $B$ is the basis for $\mathbb{R}$, formed from the columns of $P$, then $D$ is the $B$-matix for the transformation $x \mapsto Ax$.
\end{theorem}
\end{document}

>>>>

\subsection{Multiplicities of an eigenvalue}
\begin{itemize}
\item The algebraic multiplicity of $\lambda$ is the number of times $(\lambda - t)$ occurs as a factor of in its characteristic polynomial.
\item The geometric multiplicity of an eigenvalue $\lambda$ is the dimension of the corresponding eigenspace $E_{\lambda}$. i.e $dim\{E_{\lambda}\}$.
\end{itemize}
\subsection{Exercise}
Diagonalizable the following matrix, if possible.
\begin{multicols}{2}
1. $A=
\begin{bmatrix}
  4 & 0 & 0  \\
  1 & 4 & 0  \\
   0& 0 & 5  \\
\end{bmatrix}
$
\columnbreak

2. $
\begin{bmatrix}
  4 & -9 \\
  4 & 8
\end{bmatrix}
$

@@

\section{Linear Transformation and Diagonalization}
We will see that the transformation $x \to  Ax$ is essentially same as the very simple mapping $x \to Dx$ when viewed from the proper perspective.
\begin{theorem}
  Suppose $A = PDP^{-1}$, where $D$ is a diagonal $n \times n$. If $B$ is the basis for $\mathbb{R}$, formed from the columns of $P$, then $D$ is the $\mathcal{B}$-matix for the transformation $x \mapsto Ax$.
\end{theorem}
\begin{proof}
  Let the order of matrix $A$ be $n$. Since $A$ is diagonalizable, $A$ has $n$ linearly independent eigenvectors: $\mathcal{B}=\{v_1, ..., v_n\}$, that forms a basis of $\mathbb{R}^n$.
  Suppose for $x=x_1e_1 + ... + x_ne_n$ we have $x = x'_1v_1 + ...+ x'_nv_n$.
  \begin{equation}
    \label{diaglinear}
    \begin{aligned}
      Ax = A(x'_1v_1 + ...+ x'_nv_n) &= x'_1Av_1 + ... +x'_nAv_n \\[1mm]
                                      &= [Av_1 \; ... \; Av_n](x'_1 \; ... \; x'_n)^t \\[1mm]
                                      &=[\lambda_1 v_1 \; ... \; \lambda_n v_n](x'_1 \; ... \; x'_n)^t \\[1mm]
      =& \begin{bmatrix}
  \lambda_1 & 0 & ... & 0 \\
  0 & \lambda_2 & ... & 0 \\
  ... \\
  0 & 0 & ... & \lambda_1 \\
      \end{bmatrix}x \\[1mm]
      &= Dx
    \end{aligned}
  \end{equation}
  Here, $\displaystyle [\lambda_1 v_1 \; ... \; \lambda_n v_n] = \begin{bmatrix}
  \lambda_1 & 0 & ... & 0 \\
  0 & \lambda_2 & ... & 0 \\
  ... \\
  0 & 0 & ... & \lambda_1 \\
\end{bmatrix}$. This is because for $(x'_1 \; ... \; x'_n)^t$, the vectors $(v_1, ..., v_n)$ are the unit vectors, just like $e_1, ..., e_n$ are for $x_1, ..., x_n$.
\end{proof}
\begin{remark}
  \textit{This proof is not need for the students.}
\end{remark}

\subsection{Exercise}
Find a basis $\mathcal{B}$ for $\mathbb{R}^2$ with the property that the transformation matrix of the transformation given by the following matrix is Diagonal.
\begin{multicols}{2}
  1. $
  \begin{bmatrix}
    0 & 1 \\
    -3 & 4
  \end{bmatrix}
  $
  \columnbreak

  2. $
  \begin{bmatrix}
    5 & -3 \\
    -7 & 1
  \end{bmatrix}
  $

\end{multicols}
\end{document}

### 2025-06-22 lecturenotes_aima104/ch4_aima104.tex
innerprodut

>>>>

\begin{definition}
  An orthogonal basis for a vector space $V$ is a basis for $V$ that is an orthogonal set.
\end{definition}

\begin{theorem}
  \label{orthopro}
  Let $\{u_1, ..., u_p\}$ be an orthogonal basis for a subspace $W$ of $\mathbb{R}^n$. For each $y$ in $W$, the coordinates of $y$ with respect to the orthogonal basis : $y = c_1u_1 + \; ... \; c_n u_n$ are given by $\displaystyle c_j = \frac{y.u_j}{u_j.u_j}$.
\end{theorem}

\subsection{Orthogonal Projection}
The coordinate $c_j$ of $y$ in Theorem~\ref{orthopro} is actually orthogonal projection of $y$ into the vector $u_j$. This can be generalized. For any given vector $u$. The orthogonal projection of a vector $y$ on $u$ is given by the formula $\displaystyle \hat{y} = \frac{y.u}{u.u}$

Or, it can be derived as follows using the inner-product. $(y-\alpha u)$ and $u$ are orthogonal so, $(y-\alpha u).u=0$. This gives us $\displaystyle \alpha = \frac{y.u}{u.u}$. For two dimensional vectors, another orthogonal component $z$ can be easily obtained as by subtracting the projection from the vector $y$. $z = y - \hat{y}$.

\section{Orthonormal Sets}
A set $\{u_1, ..., u_p\}$ is an orthonormal set if it is an orthogonal set of unit vectors. And a basis of orthonormal set is a orthonormal basis. The simplest orthonormal basis is $\{e_1, \; ...\; e_n$ for $\mathbb{R}^n$.

Matrices whose columns form an orthonormal set are important in applications and in computer algorithms for matrix computations.

\begin{theorem}
  An $m \times n$ matrix $U$ has orthonormal columns if and only if $U^tU=I$
\end{theorem}
